% NeurIPS 2026 Paper Checklist
% Adapted from the official NeurIPS checklist

\section*{NeurIPS Paper Checklist}

\begin{enumerate}

\item {\bf Claims}
\begin{itemize}
    \item[$\checkmark$] All main claims in the abstract and introduction are supported by experiments (Sections 4-6).
    \item[$\checkmark$] Limitations are explicitly discussed (Section 7 of the full paper).
\end{itemize}

\item {\bf Theory}
\begin{itemize}
    \item[$\checkmark$] Sen's meta-ranking formalization (Eq. 1-2) includes all assumptions.
    \item[$\checkmark$] Convergence verified via ADF test (Section 5.1, 87\% of runs).
    \item[$\checkmark$] ESS convergence demonstrated via replicator dynamics (Section 6.2).
\end{itemize}

\item {\bf Experiments}
\begin{itemize}
    \item[$\checkmark$] Three environments tested: Cleanup (20/100 agents), IPD (2 agents), PGG (4 agents).
    \item[$\checkmark$] 7 SVO conditions $\times$ 5-10 seeds per condition.
    \item[$\checkmark$] All error bars represent standard deviation across seeds.
    \item[$\checkmark$] Statistical tests specified: t-test, LMM, ADF, Levene's, Cohen's d.
    \item[$\checkmark$] Sensitivity analysis across all parameter conditions (Section 5.3).
\end{itemize}

\item {\bf Reproducibility}
\begin{itemize}
    \item[$\checkmark$] Code available at \url{https://github.com/Yesol-Pilot/EthicaAI}.
    \item[$\checkmark$] Docker container provided for exact reproduction.
    \item[$\checkmark$] One-click reproduction script: \texttt{reproduce.sh}.
    \item[$\checkmark$] All hyperparameters specified in \texttt{simulation/jax/config.py}.
    \item[$\checkmark$] Random seeds explicitly set and documented.
\end{itemize}

\item {\bf Broader Impact}
\begin{itemize}
    \item[$\checkmark$] Positive: Framework for more human-like moral reasoning in AI.
    \item[$\checkmark$] Risk: Meta-ranking could be misused to create manipulative agents.
    \item[$\checkmark$] Mitigation: Dynamic $\lambda$ makes moral behavior transparent and auditable.
\end{itemize}

\item {\bf Safeguards}
\begin{itemize}
    \item[$\checkmark$] No human subjects involved.
    \item[$\checkmark$] Human PGG data from published literature (Fehr \& G\"achter, 2000).
    \item[$\checkmark$] No personally identifiable information in any dataset.
\end{itemize}

\end{enumerate}
